Speeding up decimal multiplication

### 2020-11-22 paper/paper.tex
Update

@@

Unfortunately, this method is very slow; in our benchmark, native division on x86-64
was 720\% slower than the Montgomery reduction. The code can be found in the
\verb!bench-modmul! subdirectory of our repository.

@@

We found that on a modern x86-64 system, reduction with two rounds of $r$ is slower by
$\approx$ 15\% compared to Montgomery reduction;
the code can be found in the \verb!bench-modmul! subdirectory of our repository.

@@

@@

\subsubsection{Bit wizardry}

@@

\subsubsection{Conditional moves}

@@

\begin{enumerate}
  \item \textbf{Improve cache utilization}.
    \begin{enumerate}
      \item Use breadth-first ordering, perhaps using explicit recursion.
            This can be beneficial even considering the function call overhead,
            as it leads to better cache utilization~\cite{fftw}.
      \item Explore the ``Belgian approach''; see~\cite{belgian}.
      \item Explore using higher-radix transforms (raidx-4, radix-8, etc).
            Note that radix-4 is beneficial for complex FFT because,
            for a standard representation by a pair of floating-point values,
            multiplication by $i$ or $-i$ can be done significantly faster than normal
            multiplication; but finite fields of integers modulo $p$ with standard representation
            lack this property.
            Nevertheless, higher-radix transforms might be of service because they have better cache
            locality~\cite{fftw}.
    \end{enumerate}
  \item \textbf{Additional shenanigans}.
    \begin{enumerate}
      \item It is suggested in \cite{shenanigans, mont_sp} to use primes $p$ such that $p < \mu/4.$
          Then, the Montgomery multiplication can be implemented in such a way that if the inputs
          are in $[0; 2p-1],$ then the outputs are also in $[0; 2p-1],$ and no final conditional
          subtraction is performed.
      \item It is stated in~\cite{shenanigans_presentation} that we could benefit from
          higher-radix transforms by further eliminating reductions from the butterflies.
      \item Explore signed Montgomery reduction, as defined in~\cite{signed_montgomery}.
    \end{enumerate}
  \item \textbf{Smooth the stairs}, possibly utilizing ideas from~\cite{smooth_the_stairs}.
        Currently, we can only perform transforms of lengths $2^k$ and $3 \cdot 2^k.$
        It would be nice to be able to perform transforms of lengths that are in-between.
\end{enumerate}

@@

\begin{figure}[H]
    \centering
    \begin{gnuplot}[terminal=epslatex, terminaloptions=color]
        set style line 1 lt 1 lw 2 lc rgb '#4682b4' pt -1
        set style line 2 lt 1 lw 2 lc rgb '#ee0000' pt -1

@@

        set xlabel "$log_2(n)$"
        set ylabel "ratio"

@@

        plot \
            'benchdata.txt' using 1:2 with lines ls 1 ti "CMOVs on", \
            'benchdata.txt' using 1:3 with lines ls 2 ti "CMOVs off"
    \end{gnuplot}
\end{figure}

>>>>

% ``Speeding up decimal multiplication'' (c) by shdown
%
% ``Speeding up decimal multiplication'' is licensed under a
% Creative Commons Attribution 4.0 International License.
%
% You should have received a copy of the license along with this
% work. If not, see <http://creativecommons.org/licenses/by/4.0/>.

@@

Unfortunately, this method is very slow; see section~\ref{section:modmul_bench} for the
results of our benchmark against Montgomery reduction and Solinas reduction.

@@

We found that on a modern x86-64 system, reduction with two rounds of $r$ is slower by
$\approx$10\% compared to Montgomery reduction;
see section~\ref{section:modmul_bench}.

@@

\subsubsection{Benchmark}
\label{section:modmul_bench}

We benchmarked naïve division using x86-64
\verb~divq~
instruction, Montgomery reduction,
and Solinas reduction.

Given an argument $n,$ our benchmark performs $n$ iterations of loop with
$10$ unrolled modular multiplications, performing in total $10 n$ modular multiplications.

We compiled it using Clang 11.0.0, with option \verb!-O3!. The machine is
Xiaomi RedmiBook 14" 2019 JYU4203CN laptop with
Intel\textsuperscript{\textregistered} Core\texttrademark{} i3-8145U CPU @ 2.10GHz;
CPU scaling governors for all CPUs were set to ``performance''.

\begin{center}
\begin{tabular}{ |c|c|c| }
    \hline
        Approach & $n$ & Time, s \\
    \hline\hline
        Naïve & $10^8$ & 23.04 \\
    \hline
        Montgomery & $10^8$ & 2.86 \\
    \hline
        Solinas & $10^8$ & 3.19 \\
    \hline
\end{tabular}
\end{center}

The code can be found in the \verb!bench-modmul! subdirectory of our repository
(see section~\ref{section:avail}).

@@

\subsubsection{Bit wizardry}
\label{section:bitwiz}

@@

\subsubsection{Conditional moves}
\label{section:cmovs}

@@

\begin{enumerate}
  \item \textbf{Improve cache utilization}.
    \begin{enumerate}
      \item Use breadth-first ordering, perhaps using explicit recursion.
            This can be beneficial even considering the function call overhead,
            as it leads to better cache utilization~\cite{fftw}.
      \item Explore what~\cite{gmp_paper} call the ``Belgian approach''; see~\cite{belgian}.
      \item Explore using higher-radix transforms (raidx-4, radix-8, etc).
            Note that radix-4 is beneficial for complex FFT because,
            for a standard representation by a pair of floating-point values,
            multiplication by $i$ or $-i$ can be done significantly faster than normal
            multiplication; but finite fields of integers modulo $p$ with standard representation
            lack this property.
            Nevertheless, higher-radix transforms might be of service because they have better cache
            locality~\cite{fftw}.
    \end{enumerate}
  \item \textbf{Additional shenanigans}.
    \begin{enumerate}
      \item It is suggested in \cite{shenanigans, mont_sp} to use primes $p$ such that $p < \mu/4.$
          Then, the Montgomery multiplication can be implemented in such a way that if the inputs
          are in $[0; 2p-1],$ then the outputs are also in $[0; 2p-1],$ and no final conditional
          subtraction is performed.
      \item It is stated in~\cite{shenanigans_presentation} that we could benefit from
          higher-radix transforms by further eliminating reductions from the butterflies.
      \item Explore signed Montgomery reduction, as defined in~\cite{signed_montgomery}.
    \end{enumerate}
  \item \textbf{Smooth the stairs}, possibly utilizing ideas from~\cite{smooth_the_stairs}.
        Currently, we can only perform transforms of lengths $2^k$ and $3 \cdot 2^k.$
        It would be nice to be able to perform transforms of lengths that are in-between.
\end{enumerate}

@@

We benchmarked our implementation (see section~\ref{section:avail} for the link to the repository)
against \textbf{mpdecimal}.

Since both rely on Cooley-Tukey,
the graphs of time as a function of input size $n$ (in decimal digits)
would be staircase-like.
Define ``threshold input size'' for some fixed implementation as any value of $n$ where the
discontinuity occurs --- a new ``stair'' pops up.

We calculated the threshold input sizes of both our implementation and of the \textbf{mpdecimal}
library, within reasonable limits ($2176 \le n \le 3\cdot10^7$).
We merged all threshold values into a single list, sorted it, then formed a new list from the
averages of each two adjacent values in the sorted list.
We then appended the maximum threshold value plus one to the resulting list.
It is easy to see that this way nothing gets lost --- assuming that the staircase, aside from
discontinuities, is horizontal, we measure all possible cases.

For each value of $n,$ we performed $\lfloor \frac{8 \cdot 10^7}{n} \rfloor$
multiplications of numbers of $n$ decimal digits each.

We compiled everything using Clang 11.0.0, with option \verb!-O3!. The machine is
Xiaomi RedmiBook 14" 2019 JYU4203CN laptop with
Intel\textsuperscript{\textregistered} Core\texttrademark{} i3-8145U CPU @ 2.10GHz;
CPU scaling governors for all CPUs were set to ``performance''.

We used \textbf{mpdecimal} version 2.5.0-4 from Debian Bullseye repositories
for the ``amd64'' architecture.
For further information on \textbf{mpdecimal}, refer to
its homepage \url{http://www.bytereef.org/mpdecimal/index.html}.

\begin{figure}[H]
    \centering
    \begin{gnuplot}[terminal=epslatex, terminaloptions=color]

        set style line 1 lt 1 lw 2 lc rgb '#4682b4' pt -1
        set style line 2 lt 1 lw 2 lc rgb '#ee0000' pt -1

@@

        set xlabel "$log_2(n)$"
        set ylabel "Ratio"

@@

        plot \
            'benchdata.txt' using 1:2 with lines ls 1 ti "CMOVs on", \
            'benchdata.txt' using 1:3 with lines ls 2 ti "CMOVs off"

    \end{gnuplot}
\end{figure}

``Ratio'' is time of \textbf{mpdecimal} divided by time of our implementation.
The ``CMOVs on'' variant uses Clang-specific
\verb~__builtin_unpredictable~ built-in for modular addition, subtraction, and adjustment
(see section~\ref{section:cmovs});
the ``CMOVs off'' variant relies on ``bit wizardry'' instead (see section~\ref{section:bitwiz}).

\section{Availability}
\label{section:avail}

The code of our implementation, benchmark scripts, and \LaTeX source of this paper,
are available at \url{https://github.com/shdown/decimal-multiplication-paper}.
The code is licensed under the MIT license.
The source of this paper is licensed under the Creative Commons BY 4.0 license.
